---
title: "Topographic Change caused by People"
author: "Matthew Ross (Kathryn Willi HW Submission)"
date: "9/23/2019"
output: html_document
---


# Analyzing topographic change in the USA

```{r setup, include=FALSE}
library(tidyverse)
library(sf)  #Spatial datasets
library(mapview) #interactive mapping
library(raster) #Raster analysis
library(elevatr) #National Elevation Dataset downloader. 

knitr::opts_chunk$set(echo = TRUE)
```


## USGS analysis of topographic change

Analysis background can be found here: https://www.usgs.gov/land-resources/eros/topochange


### Download the data


```{r, eval=F}
#Create a new folder 
dir.create('data')

#Download files
download.file(url='https://edcintl.cr.usgs.gov/downloads/sciweb1/shared/topo/downloads/Topochange/topo_change_polygons_Shapefile.zip',destfile='data/topo_change.zip')

# List files in the data directory
#list.files('data')

#Unzip the dataset
unzip('data/topo_change.zip',exdir='data/topo_change')

#Check that the unzipping worked
#list.files('data/topo_change')

```

### Check out the data

```{r}

topo_change <- st_read('data/topo_change/topo_change_polygons.shp')

#Checkout the column names
#names(topo_change)

biggest_area <- topo_change %>%
  arrange(desc(AREA_SQ_KM)) %>%
  slice(1:10)

# Map all mines
mapview(biggest_area)
```

### Checkout mine in Arizona

```{r}
# Subset to just Arizona
az_mine <- biggest_area %>%
  filter(QUADNAME == 'esperanza_mill_AZ')

#Check that it is the right site
mapview(az_mine)
```



## Getting our own elevation datasets

### Download pre-mining DEM 

```{r}

#Check projection of az_mine
st_crs(az_mine)

az_raster_before <- get_elev_raster(az_mine,z=12)

#Look at the structure of the data
#str(az_raster_before)

#Summary of the data
#summary(az_raster)

plot(az_raster_before)

```


### Download post-mining DEM

Data comes from [SRTM](https://www2.jpl.nasa.gov/srtm/) 

```{r,eval=F}
#Extract centroid of mine polygon
az_center <- st_centroid(az_mine) %>%
  st_transform(4326)
#Map it to check
mapview(az_center)

#Get coords 
coords <- st_coordinates(az_center)


getData('SRTM',lon=coords[1],lat=coords[2],
        path='data')

```


### Read in SRTM data
```{r}
az_raster_after <- raster('data/srtm_14_06.tif')

```

## Looking at Elevation change before and after mining

### Clip SRTM data to NED data

```{r}
#Check resolution of rasters
res(az_raster_after)
res(az_raster_before)

#Check the projection
projection(az_raster_after)
projection(az_raster_before)


#Reproject small raster to larger raster projection
az_raster_before_reproj <- projectRaster(az_raster_before,
                                        az_raster_after)

#Trim whitespace
az_raster_before_trim <- trim(az_raster_before_reproj) 

#Crop to match extents
az_raster_after_clip <- crop(az_raster_after,az_raster_before_trim)

#Plot clipped dataset
plot(az_raster_after_clip)
```


### Taking the difference and plotting it

```{r}
# Take the elevation difference
dem_diff <- az_raster_before_trim - az_raster_after_clip

plot(dem_diff)

```


# In-class work

## 1) What is the maximum/minimum elevation change caused by mining? You may want to check out the command maxValue and minValue


```{r}

#YOUR CODE HERE

maxValue(dem_diff)

minValue(dem_diff)
```


## 2) What is the background elevation "error" between NED and SRTM? 

The national elevation dataset (NED) and the Shuttle 
Radar Topography Mission are extremely different ways to
estimate elevatino. NED is derived from thousands of 
compiled elevation datasets that can be collected on the
ground, from LiDAR or other ways, while SRTM is collected
with radar from space. As a result, we naturally expect
there to be some background differences in these 
two DEMs, so what is the average difference in these DEMs
in parts of the landscape that have not undergone substantial 
elevation change? 

### 2a) Buffer the az_mine object by 5 kilometers using the sf::st_buffer command

```{r}

#Your code here
az_mine_buffer <- sf::st_buffer(az_mine, 5000)

mapview(az_mine_buffer)
```


### 2b) Use the the st_transform command to transform the projection of your buffered az_mine dataset to the same projection as the dem_diff raster dataset

hint: you may want to explore st_crs to help you here

```{r}

projection(dem_diff)

st_crs(az_mine_buffer)

az_mine_buffer_reproj <- st_transform(az_mine_buffer, 4326)

projection(az_mine_buffer_reproj)

```


### 2c) Use the raster::mask command to mask out all the  areas inside of the buffered and reprojected az_mine dataset.

hint: look at the mask command and try to understand what
the inverse option does

```{r}
#Your code here

unmined_dem_diff <- raster::mask(dem_diff,az_mine_buffer_reproj, inverse=TRUE)
mapview(unmined_dem_diff)

```


### 2d) Use the summary command to look at the summary of differences between these two DEMs in areas where there has been no mining

```{r}

summary(unmined_dem_diff)

```

*Based on the Q1 and Q3 interquartile range, the differences aren't TOO substantial, but still exist. However there are some major outliers!* 

## 3) What does the shape of the "error" distribution look like in areas that have not been mined versus areas that have been mined? 

To look at distributional changes, we may want to convert this
raster dataset back into a data.frame or tibble type shape so 
that we can take advantage of ggplots plotting capabilities and 
make a plot of the two dems.

### 3a) Mask the raster dem_diff dataset to the exact outlines of the USGS az_mine dataset

```{r}

az_mine_wgs84 <- st_transform(az_mine, 4326)

projection(az_mine_wgs84)

mined_dem_diff <- raster::mask(dem_diff, az_mine_wgs84)
mapview(mined_dem_diff)

```


### 3b) Use the getValues command to extract the elevation values from both the unmined and mined regions of your rasters.  Label these vectors as `unmined_dem` and `mined_dem`. 

```{r}
unmined_dem <- getValues(unmined_dem_diff)
mined_dem <- getValues(mined_dem_diff)


```


### 3c) Use ggplot and the command geom_density() to look at the differences in mined and unmined regions of Arizona.

```{r}
# Here I'm just making the table for you

unmined_type <- rep('unmined',length(unmined_dem))
mined_type <- rep('mined',length(mined_dem))
mined_table <- tibble(dem_values = c(unmined_dem,mined_dem),
                      type = c(unmined_type,mined_type)) 
                    

ggplot(mined_table, aes(dem_values, fill=type, color=type)) +
  geom_density() 
  
                    
```

*As one would likely expect, there are many 'minor' differences between the two rasters in the unmined region, while there are substantial differences between the two rasters in the mined region.*

